Distance Sampling in \proglang{R}

### 2016-06-08 paper.tex
fix-up
got through some of the requirements

\documentclass[article]{jss}
\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
David L Miller\\Integrated Statistics and University of St Andrews \AND Eric Rexstad\\University of St Andrews \And Len Thomas\\University of St Andrews \AND Laura Marshall\\University of St Andrews \And Jeffrey L Laake\\Marine Mammal Laboratory
}
\title{Distance Sampling in \proglang{R}}
\Keywords{distance sampling, abundance estimation, line transect, point transect, detection function, Horvitz-Thompson, \proglang{R}, Distance}

\Abstract{
Estimating the abundance and spatial distribution of animal and plant
populations is essential for conservation and management. We introduce
the \proglang{R} package \pkg{Distance} that implements distance
sampling methods to estimate abundance. We describe how users can obtain
estimates of abundance (and density) using the package as well
documenting the links it provides with other more specialized
\proglang{R} packages. We also demonstrate how \pkg{Distance} provides a
migration pathway from previous software, thereby allowing us to deliver
cutting-edge methods to the users more quickly.
}

\Plainauthor{David L Miller, Eric Rexstad, Len Thomas, Laura Marshall, Jeffrey L Laake}
\Plaintitle{Distance Sampling in R}
\Plainkeywords{distance sampling, abundance estimation, line transect, point transecs, detection function, Horvitz-Thompson, R, Distance}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
\Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    David L Miller\\
  Integrated Statistics and University of St Andrews\\
  16 Sumner Street, Woods Hole, MA 02543, USA and Centre for Research into
  Ecological and Environmental Modelling, The Observatory, St Andrews,
  Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:dave@ninepointeightone.net}{\nolinkurl{dave@ninepointeightone.net}}\\
  URL: \url{http://converged.yt}\\~\\
      Eric Rexstad\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling, The
  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:Eric.Rexstad@st-andrews.ac.uk}{\nolinkurl{Eric.Rexstad@st-andrews.ac.uk}}\\

      Len Thomas\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling, The
  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:len.thomas@st-andrews.ac.uk}{\nolinkurl{len.thomas@st-andrews.ac.uk}}\\
  URL: \url{http://lenthomas.org}\\~\\
      Laura Marshall\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling, The
  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:lhm@st-andrews.ac.uk}{\nolinkurl{lhm@st-andrews.ac.uk}}\\

      Jeffrey L Laake\\
  Marine Mammal Laboratory\\
  Alaska Fisheries Science Center National Oceanic and Atmospheric
  Administration 7600 Sand Point Way N.E., Seattle, WA 98115, USA\\
  E-mail: \href{mailto:Jeff.Laake@noaa.gov}{\nolinkurl{Jeff.Laake@noaa.gov}}\\

  }

\usepackage{amsmath} \usepackage{amssymb} \usepackage{longtable}
\usepackage{booktabs}
\usepackage[colorinlistoftodos, color=green!40, textwidth=60, textsize=small]{todonotes}

@@

\section{Introduction}\label{introduction}

@@

Conventional distance sampling assumes the observer is located either at
a point or moving along a line and will observe all objects that occur
at the point or on the line. The further away an object is from the
point or line (also known as the sampler or transect) the less likely it
is that the observer will see it. We can use the distances to each of
the detected objects from the line or point to build a model of the
probability of detection given distance from the sampler -- the
\emph{detection function}. The detection function can be used to infer
how many objects were missed and thereby produce estimates of density
and/or abundance.

The Windows program Distance (or ``DISTANCE''; for clarity henceforth
``Distance for Windows'') can be used to fit detection functions to
distance sampling data. It was first released (versions 1.0 - 3.0;
principally programmed by Jeff Laake while working at the National
Marine Mammal Laboratory) as a console-based application (this in turn
was based on earlier software TRANSECT, \citet{Burnham:1980wz} and
algorithms developed in \citet{Buckland:1992fa}), before the first
graphical interface (Distance for Windows 3.5) was released in November
1998. Since this time it has evolved to include various design and
analysis features \citep{Thomas:2010cf}. Distance for Windows versions 5
onwards have included \proglang{R} \citep{rcore} packages as the
analysis engines providing additional, more complex analysis options
than those offered by the original (FORTRAN) code.

@@

\subsection{Distance sampling}\label{distance-sampling}

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/detectdists-1}

}

\caption{Panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row). The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling. In this example, detection probability becomes effectively zero at 500 (distances shown on $x$-axis are arbitrary). \label{fig:detectiondist}}\label{fig:detectdists}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
Warning: package 'nlme' was built under R version 3.2.4
\end{CodeOutput}
\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/points-and-lines-1}

}

\caption{Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals. The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\label{fig:pointslines}}\label{fig:points-and-lines}
\end{figure}
\end{CodeChunk}

Good survey design is essential to produce reliable density and
abundance estimates from distance sampling data. Survey design is beyond
the scope of this article but we refer readers to \citet[Chapter
7]{Buckland:2001vm} and \citet[Chapter 2]{buckland2015distance} for
introductory information; \citet{Strindberg:2004vr} contains information
on automated survey design using geographical information systems (GIS);
\citet{Thomas:2007wz} gives an example of designing a distance sampling
survey in a complex region. We also note that Distance for Windows
includes a GIS survey design tool \citep{Thomas:2010cf}.

@@

Having estimated the detection function's parameters, one can then
integrate out distance from the function (as the detection function
describes the probability of detection \emph{given} distance) to get an
``average'' probability of detection, which can be used to correct the
observed counts. Summing the corrected counts gives an estimate of
abundance in the area covered by surveys, which can be multiplied up to
the total study area.

In addition to randomly placed samplers distance sampling relies on
three other main assumptions. Firstly, all objects directly on the
transect line or point (i.e.~those at zero distance) are detected (see
``Extensions'' for methods to deal with the situation when this is not
possible). Secondly, objects are stationary or detected prior to any
movement. Thirdly, distance to the object must be measured accurately,
or the observation allocated to the correct distance bin for grouped
data. Depending on the survey species some of these assumptions may be
more difficult to meet than others. Further information on field methods
to help meet these assumptions can be found in \citet{Buckland:2001vm}
and \citet{buckland2015distance}.

@@

\section{Data}\label{data}

We introduce two example analyses performed in \pkg{Distance}: one line
transect and one point transect. These data sets have been chosen as
they represent typical data seen in practice. The below example analyses
are not intended to serve as guidelines, but to demonstrate features of
the software. Practical advice on approaches to analysis is given in
\citet{Thomas:2010cf}.

\subsubsection{Minke whales}\label{minke-whales}

The line transect data have been simulated from models fitted to
Antarctic minke whale (\emph{Balaenoptera bonaerensis}) data. These data
were collected as part of the International Whaling Commission's
International Decade of Cetacean Research Southern Ocean Whale and
Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer
surveys. These data consist of 99 observations that are stratified based
on location (near or distant from ice edge) and effort data (transect
lengths). Further details on the survey are available in
\citet{Branch:2001ua} (data is simulated based on the design used for
``1992/93 Area III'' therein).

\subsubsection{Amakihi}\label{amakihi}

The point transect data set consists of 1485 observations of Amakihi
(\emph{Hemignathus virens}; a Hawaiian songbird), collected at 41 points
between 1992 and 1995. The data include distances and two covariates
collected during the survey: observer (a three level factor), time after
sunrise (transformed to minutes (continuous) or hours (factor)
covariates). Data are analysed comprehensively in
\citet{Marques:2007ey}.

\subsection{Data setup}\label{data-setup}

@@

\begin{itemize}
\tightlist
\item
  In the simplest case, where the objective is to estimate a detection
  function and exact distances are collected, all that is required is a
  numeric vector of distances.
\item
  To include additional covariates into the detection function (see
  ``Detection functions'') a \code{data.frame} is required. Each row of
  the \code{data.frame} contains the data on one observation. The
  \code{data.frame} must contain a column named \code{distance}
  (containing the observed distances) and additional named columns for
  any covariates that may affect detectability (for example
  \code{observer} or \code{seastate}). The column name \code{size} is
  reserved for the cluster sizes (sometimes referred to as group sizes)
  in this case each row represents an observation of a cluster rather
  than individual (see \citet{Buckland:2001vm}, Section 1.4.3 for more
  on defining clusters and ``Extensions'' below for one approach to
  dealing with uncertain cluster size). Additional reserved names
  include \code{object} and \code{detected}, these are not required for
  conventional distance sampling and should be avoided (see
  ``Extensions'' for an explanation of their use).
\item
  To estimate density or to estimate abundance beyond the sampled area,
  additional information is required. Additional columns should be
  included in the \code{data.frame} specifying: \code{Sample.Label}, the
  ID of the transect; \code{Effort}, transect effort (for lines their
  length and for points the number of times that point was visited);
  \code{Region.Label}, the stratum containing the transect (which may be
  from pre- or post-survey stratification, see ``Estimating abundance
  and variance''); \code{Area}, the area of the strata. Transects which
  were surveyed but have no observations must be included in the data
  set with \code{NA} for \code{distance} and any other covariates. We
  refer to this data format (where all information is contained in one
  table) as ``flatfile'' as it can be easily created in a single
  spreadsheet.
\end{itemize}

@@

It is also possible to use distances collected in intervals (``binned''
or ``grouped'' data, as opposed to ``exact'' distances) to model the
detection function. In this case the column \code{distance} is replaced
by two columns \code{distbegin} and \code{distend} referring to the
distance interval start and end cutpoints. More information on binned
data is included in \citet{Buckland:2001vm} sections 4.5 and 7.4.1.2.

@@

The minke whale data follows the ``flatfile'' format given in the last
bullet point:

\begin{CodeChunk}
\begin{CodeInput}
library("Distance")
head(minke)
\end{CodeInput}
\begin{CodeOutput}
  Region.Label  Area Sample.Label Effort distance
1        South 84734            1  86.75     0.10
2        South 84734            1  86.75     0.22
3        South 84734            1  86.75     0.16
4        South 84734            1  86.75     0.78
5        South 84734            1  86.75     0.21
6        South 84734            1  86.75     0.95
\end{CodeOutput}
\end{CodeChunk}

Whereas the amakihi data lacks effort and stratum data:

\begin{CodeChunk}
\begin{CodeInput}
head(amakihi)
\end{CodeInput}
\begin{CodeOutput}
   survey object distance obs mas has detected
1 July 92      1       40 TJS  50   1        1
2 July 92      2       60 TJS  50   1        1
3 July 92      3       45 TJS  50   1        1
4 July 92      4      100 TJS  50   1        1
5 July 92      5      125 TJS  50   1        1
6 July 92      6      120 TJS  50   1        1
\end{CodeOutput}
\end{CodeChunk}

@@

\section{Detection functions}\label{detection-functions}

The detection function models the probability
\(\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)\)
and is usually denoted \(g(y; \boldsymbol{\theta})\) where \(y\) is
distance (from a line or point) and \(\boldsymbol{\theta}\) is a vector
of parameters to be estimated. Our goal is to estimate an \emph{average
probability of detection} (\(p\), average in the sense of an average
over distance from \(0\) to truncation distance \(w\)), so we must
integrate out distance (\(y\)) from the detection function: \[
p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
\] where \(\pi(y)\) describes the distribution of objects with respect
to the sampler; \(\pi(x)=1/w\) for line transects and
\(\pi(r)=\frac{2r}{w^2}\) for point transects, taking into account the
geometry of the sampler \citep[usually referred to as the
\emph{probability density function of distances};][Chapter
3]{Buckland:2001vm}. When considering a particular transect type we let
\(x\) denote a perpendicular distance from a line and \(r\) denote
radial distance from a point (rather than using \(y\)).

It is crucial that the detection function accurately models
detectability at small distances; we are less worried by its behaviour
further away from 0. To ensure that the model is not overly influenced
by distances far from zero, we truncate the distances beyond a given
distance \(w\) (this can be done during analysis or while collecting
data in the field).

Models for the detection function are expected to have the following
properties \citep[Chapter 5]{buckland2015distance}:

\begin{itemize}
\tightlist
\item
  \emph{Shoulder}: we expect observers to be able to see objects near
  them, not just those directly in front of them. For this reason, we
  expect the detection function to be flat near the line or point.
\item
  \emph{Non-increasing}: we do not think that observers should be more
  likely to see distant objects than those nearer the transect. If this
  occurs, it usually indicates an issue with survey design or field
  procedures (for example that the distribution of objects with respect
  to the line, \(\pi(y)\) is not what we expect), so we do not want the
  detection function to model this
  \citep{Marques:2010he, Marques:2012fy, Miller:2015hw}.
\item
  \emph{Model robust}: models should be flexible enough to fit many
  different shapes.
\item
  \emph{Pooling robust}: many factors can affect the probability of
  detection and it is not possible to measure all of these. We would
  like models to produce unbiased results without inclusion of these
  factors.
\item
  \emph{Estimator efficiency}: we would like models to have low
  variances, given they satisfy the other properties above (which, if
  satisfied, would give low bias).
\end{itemize}

Given these criteria, we can formulate models for \(g\).

\subsection{Formulations}\label{formulations}

@@

\subsubsection{Key function plus adjustments
(K+A)}\label{key-function-plus-adjustments-ka}

Key function plus adjustment terms (or adjustment series) models are
formulated by taking a ``key'' function and optionally adding
``adjustments'' to it to improve the fit \citep{Buckland:1992fa}.
Mathematically we formulate this as: \[
g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
\] where \(k\) is the key function and \(\alpha_O\) is sum series of
functions (given in Table \ref{tab:keyadj}), described as an
\emph{adjustment of order \(O\)}. Subscripts on the parameter vector
indicate those parameters belonging to each part of the model (i.e.
\(\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})\)).

Available models for the key are as follows: \[
k(y) = \left\{
\begin{array}{l l}
  \exp\left(-\frac{y^2}{2 \sigma^2}\right) & \quad \text{half-normal,} \\
  1-\exp\left(\left(-\frac{y}{\sigma}\right)^{-b}\right) & \quad \text{hazard-rate,} \\
  1/w & \quad \text{uniform.}
\end{array} \right.
\] Possible modelling options for key and adjustments are given in Table
\ref{tab:keyadj} and illustrated in Figures \ref{fig:hnhr} and
\ref{fig:keyadj}. We select the number of adjustment terms (\(K\)) by
AIC (further details in ``Model checking and model selection'').

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/hn-hr-par-comp-1}

}

\caption{Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots). On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances). The bottom row shows the hazard-rate model's more pronounced shoulder.\label{fig:hnhr}}\label{fig:hn-hr-par-comp}
\end{figure}
\end{CodeChunk}

\begin{table}
\begin{tabular}{llll}
\hline
Key function   & Form   & Adjustment series & Form\\
\hline
 Uniform  & $1/w$   & cosine  & $\sum_{o=1}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=1}^O a_o (y/w)^{2o}$ \\
 Half-normal  & $\exp\left(-\frac{y^2}{2 \sigma^2}\right)$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Hermite polynomial & $\sum_{o=2}^O a_o H_{2o}(y/\sigma)$ \\
 Hazard-rate  & $1-\exp\left[-\left(\frac{y}{\sigma}\right)^{-b}\right]$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=2}^O a_o (y/w)^{2o}$ \\
\hline
\end{tabular}
\label{tab:keyadj}
\caption{Modelling options for key plus adjustment series models for the detection function. For each key function the default adjustments are cosine in \pkg{Distance}. Note that in the adjustments functions distance is divided by the width or the scale parameter to ensure the shape of adjustment independent of the units of $y$ (Marques et al., 2007); defaults are shown here, though either can be selected to rescale the distances.}
\end{table}

>>>>

\documentclass[article]{jss}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{amsmath}
\usepackage{amssymb}
%% need no \usepackage{Sweave.sty}

%\usepackage{longtable}
\usepackage{booktabs}

%% almost as usual
\author{David L Miller\\Integrated Statistics and University of St Andrews \AND 
        Eric Rexstad\\University of St Andrews \And 
        Len Thomas\\University of St Andrews \AND 
        Laura Marshall\\University of St Andrews \And 
        Jeffrey L Laake\\Marine Mammal Laboratory}

\title{Distance Sampling in \proglang{R}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{David L Miller, Eric Rexstad, Len Thomas, Laura Marshall, Jeffrey L Laake}
\Plaintitle{Distance Sampling in R} %% without formatting
%\Shorttitle{\pkg{foo}: A Capitalized Title} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
Estimating the abundance and spatial distribution of animal and plant populations is essential for conservation and management. We introduce the \proglang{R} package \pkg{Distance} that implements distance sampling methods to estimate abundance. We describe how users can obtain estimates of abundance (and density) using the package as well documenting the links it provides with other more specialized \proglang{R} packages. We also demonstrate how \pkg{Distance} provides a migration pathway from previous software, thereby allowing us to deliver cutting-edge methods to the users more quickly.
}
\Keywords{distance sampling, abundance estimation, line transect, point transect, detection function, Horvitz-Thompson, \proglang{R}, Distance}
\Plainkeywords{distance sampling, abundance estimation, line transect, point transecs, detection function, Horvitz-Thompson, R, Distance} %% without formatting
%% at least one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  David L Miller\\
  Integrated Statistics and University of St Andrews\\
  16 Sumner Street, Woods Hole, MA 02543, USA and\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{dave@ninepointeightone.net}\\
  URL: \url{http://converged.yt} \\ \\
  Eric Rexstad\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{Eric.Rexstad@st-andrews.ac.uk} \\ \\
  Len Thomas\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{len.thomas@st-andrews.ac.uk}\\
  URL: \url{http://lenthomas.org} \\ \\
  Laura Marshall\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling,\\
  The Observatory,\\
  St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \email{lhm@st-andrews.ac.uk}\\ \\
  Jeffrey L Laake\\
  Marine Mammal Laboratory\\
  Alaska Fisheries Science Center\\
  National Oceanic and Atmospheric Administration\\
  7600 Sand Point Way N.E., Seattle, WA 98115, USA\\
  E-mail: \email{Jeff.Laake@noaa.gov}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section{Introduction}

@@

Conventional distance sampling assumes the observer is located either at a point or moving along a line and will observe all objects that occur at the point or on the line. The further away an object is from the point or line (also known as the sampler or transect) the less likely it is that the observer will see it. We can use the distances to each of the detected objects from the line or point to build a model of the probability of detection given distance from the sampler --- the \textit{detection function}. The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance.

The Windows program Distance (or ``DISTANCE''; for clarity henceforth ``Distance for Windows'') can be used to fit detection functions to distance sampling data. It was first released (versions 1.0 - 3.0; principally programmed by Jeff Laake while working at the National Marine Mammal Laboratory) as a console-based application (this in turn was based on earlier software TRANSECT, \citealp{Burnham:1980wz} and algorithms developed in \citealp{Buckland:1992fa}), before the first graphical interface (Distance for Windows 3.5) was released in November 1998. Since this time it has evolved to include various design and analysis features \citep{Thomas:2010cf}. Distance for Windows versions 5 onwards have included \proglang{R} \citep{rcore} packages as the analysis engines providing additional, more complex analysis options than those offered by the original (\proglang{Fortran}) code.

@@

\subsection{Distance sampling}

@@

\begin{figure}
\begin{center}
\includegraphics{paper-detectdists}
\caption{Panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row). The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling. In this example, detection probability becomes effectively zero at 500 (distances shown on $x$-axis are arbitrary).}
\label{fig:detectiondist}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics{paper-points-and-lines}
\caption{Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals. The right side of the figure shows histograms of observed distances (again, lines top and points bottom).}
\label{fig:pointslines}
\end{center}
\end{figure}

Good survey design is essential to produce reliable density and abundance estimates from distance sampling data. Survey design is beyond the scope of this article but we refer readers to \cite[][Chapter 7]{Buckland:2001vm} and \cite[][Chapter 2]{buckland2015distance} for introductory information; \cite{Strindberg:2004vr} contains information on automated survey design using geographical information systems (GIS); \cite{Thomas:2007wz} gives an example of designing a distance sampling survey in a complex region. We also note that Distance for Windows includes a GIS survey design tool \citep{Thomas:2010cf}.

@@

Having estimated the detection function's parameters, one can then integrate out distance from the function (as the detection function describes the probability of detection \textit{given} distance) to get an ``average'' probability of detection, which can be used to correct the observed counts. Summing the corrected counts gives an estimate of abundance in the area covered by surveys, which can be multiplied up to the total study area.

In addition to randomly placed samplers distance sampling relies on three other main assumptions. Firstly, all objects directly on the transect line or point (i.e., those at zero distance) are detected (see ``Extensions'' for methods to deal with the situation when this is not possible). Secondly, objects are stationary or detected prior to any movement. Thirdly, distance to the object must be measured accurately, or the observation allocated to the correct distance bin for grouped data. Depending on the survey species some of these assumptions may be more difficult to meet than others. Further information on field methods to help meet these assumptions can be found in \cite{Buckland:2001vm} and \cite{buckland2015distance}.

@@

\section{Data}

We introduce two example analyses performed in \pkg{Distance}: one line transect and one point transect. These data sets have been chosen as they represent typical data seen in practice. The below example analyses are not intended to serve as guidelines, but to demonstrate features of the software. Practical advice on approaches to analysis is given in \cite{Thomas:2010cf}.

\subsection{Minke whales}

The line transect data have been simulated from models fitted to Antarctic minke whale (\textit{Balaenoptera bonaerensis}) data. These data were collected as part of the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys. These data consist of 99 observations that are stratified based on location (near or distant from ice edge) and effort data (transect lengths). Further details on the survey are available in \cite{Branch:2001ua} (data is simulated based on the design used for ``1992/93 Area III'' therein).

\subsection{Amakihi}

The point transect data set consists of 1485 observations of Amakihi (\textit{Hemignathus virens}; a Hawaiian songbird), collected at 41 points between 1992 and 1995. The data include distances and two covariates collected during the survey: observer (a three level factor), time after sunrise (transformed to minutes (continuous) or hours (factor) covariates). Data are analysed comprehensively in \cite{Marques:2007ey}.

\subsection{Data setup}

@@

\begin{itemize}
\item In the simplest case, where the objective is to estimate a detection function and exact distances are collected, all that is required is a numeric vector of distances.
\item To include additional covariates into the detection function (see ``Detection functions'') a \code{data.frame} is required. Each row of the \code{data.frame} contains the data on one observation. The \code{data.frame} must contain a column named \code{distance} (containing the observed distances) and additional named columns for any covariates that may affect detectability (for example \code{observer} or \code{seastate}). The column name \code{size} is reserved for the cluster sizes (sometimes referred to as group sizes) in this case each row represents an observation of a cluster rather than individual \citep[see][Section 1.4.3 for more on defining clusters and ``Extensions'' below for one approach to dealing with uncertain cluster size]{Buckland:2001vm}. Additional reserved names include \code{object} and \code{detected}, these are not required for conventional distance sampling and should be avoided (see ``Extensions'' for an explanation of their use).
\item To estimate density or to estimate abundance beyond the sampled area, additional information is required. Additional columns should be included in the \code{data.frame} specifying: \code{Sample.Label}, the ID of the transect; \code{Effort}, transect effort (for lines their length and for points the number of times that point was visited); \code{Region.Label}, the stratum containing the transect (which may be from pre- or post-survey stratification, see ``Estimating abundance and variance''); \code{Area}, the area of the strata. Transects which were surveyed but have no observations must be included in the data set with \code{NA} for \code{distance} and any other covariates. We refer to this data format (where all information is contained in one table) as ``flatfile'' as it can be easily created in a single spreadsheet.
\end{itemize}

@@

It is also possible to use distances collected in intervals (``binned'' or ``grouped'' data, as opposed to ``exact'' distances) to model the detection function. In this case the column \code{distance} is replaced by two columns \code{distbegin} and \code{distend} referring to the distance interval start and end cutpoints. More information on binned data is included in \cite{Buckland:2001vm} sections 4.5 and 7.4.1.2.

@@

The minke whale data follows the ``flatfile'' format given in the last bullet point:
\begin{Schunk}
\begin{Sinput}
R> library("Distance")
R> data("minke")
R> head(minke)
\end{Sinput}
\begin{Soutput}
  Region.Label  Area Sample.Label Effort distance
1        South 84734            1  86.75     0.10
2        South 84734            1  86.75     0.22
3        South 84734            1  86.75     0.16
4        South 84734            1  86.75     0.78
5        South 84734            1  86.75     0.21
6        South 84734            1  86.75     0.95
\end{Soutput}
\end{Schunk}
Whereas the amakihi data lacks effort and stratum data:
\begin{Schunk}
\begin{Sinput}
R> data("amakihi")
R> head(amakihi)
\end{Sinput}
\begin{Soutput}
   survey object distance obs mas has detected
1 July 92      1       40 TJS  50   1        1
2 July 92      2       60 TJS  50   1        1
3 July 92      3       45 TJS  50   1        1
4 July 92      4      100 TJS  50   1        1
5 July 92      5      125 TJS  50   1        1
6 July 92      6      120 TJS  50   1        1
\end{Soutput}
\end{Schunk}

@@

\section{Detection functions}

The detection function models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated. Our goal is to estimate an \textit{average probability of detection} ($p$, average in the sense of an average over distance from $0$ to truncation distance $w$), so we must integrate out distance ($y$) from the detection function:
$$
p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
$$
where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler \citep[usually referred to as the \textit{probability density function of distances}][Chapter 3]{Buckland:2001vm}, . When considering a particular transect type we let $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point (rather than using $y$).

It is crucial that the detection function accurately models detectability at small distances; we are less worried by its behaviour further away from 0. To ensure that the model is not overly influenced by distances far from zero, we truncate the distances beyond a given distance $w$ (this can be done during analysis or while collecting data in the field).

Models for the detection function are expected to have the following properties \citep[][Chapter 5]{buckland2015distance}:

\begin{itemize}
\item \textit{Shoulder}: we expect observers to be able to see objects near them, not just those directly in front of them. For this reason, we expect the detection function to be flat near the line or point.
\item \textit{Non-increasing}: we do not think that observers should be more likely to see distant objects than those nearer the transect. If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this \citep{Marques:2010he, Marques:2012fy, Miller:2015hw}.
\item \textit{Model robust}: models should be flexible enough to fit many different shapes.
\item \textit{Pooling robust}: many factors can affect the probability of detection and it is not possible to measure all of these. We would like models to produce unbiased results without inclusion of these factors.
\item \textit{Estimator efficiency}: we would like models to have low variances, given they satisfy the other properties above (which, if satisfied, would give low bias).
\end{itemize}

Given these criteria, we can formulate models for $g$.

\subsection{Formulations}

@@

\subsubsection{Key function plus adjustments (K+A)}

Key function plus adjustment terms (or adjustment series) models are formulated by taking a ``key'' function and optionally adding ``adjustments'' to it to improve the fit \citep{Buckland:1992fa}. Mathematically we formulate this as:
$$
g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
$$
where $k$ is the key function and $\alpha_O$ is sum series of functions (given in Table \ref{tab:keyadj}), described as an \textit{adjustment of order $O$}. Subscripts on the parameter vector indicate those parameters belonging to each part of the model (i.e., $\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})$).

Available models for the key are as follows:
$$
k(y) = \left\{
\begin{array}{l l}
  \exp\left(-\frac{y^2}{2 \sigma^2}\right) & \quad \text{half-normal,} \\
  1-\exp\left(\left(-\frac{y}{\sigma}\right)^{-b}\right) & \quad \text{hazard-rate,} \\
  1/w & \quad \text{uniform.}
\end{array} \right.
$$
Possible modelling options for key and adjustments are given in Table \ref{tab:keyadj} and illustrated in Figures \ref{fig:hnhr} and \ref{fig:keyadj}. We select the number of adjustment terms ($K$) by AIC (further details in ``Model checking and model selection'').

\begin{figure}
\begin{center}
\includegraphics{paper-hn-hr-par-comp}
\caption{Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots). On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances). The bottom row shows the hazard-rate model's more pronounced shoulder.}
\label{fig:hnhr}
\end{center}
\end{figure}

\begin{table}
\caption{Modelling options for key plus adjustment series models for the detection function. For each key function the default adjustments are cosine in \pkg{Distance}. Note that in the adjustments functions distance is divided by the width or the scale parameter to ensure the shape of adjustment independent of the units of $y$ \citep{Marques:2007ey}; defaults are shown here, though either can be selected to rescale the distances.}
\begin{tabular}{llll}
\hline
Key function   & Form   & Adjustment series & Form\\
\hline
 Uniform  & $1/w$   & cosine  & $\sum_{o=1}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=1}^O a_o (y/w)^{2o}$ \\
 Half-normal  & $\exp\left(-\frac{y^2}{2 \sigma^2}\right)$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Hermite polynomial & $\sum_{o=2}^O a_o H_{2o}(y/\sigma)$ \\
 Hazard-rate  & $1-\exp\left[-\left(\frac{y}{\sigma}\right)^{-b}\right]$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=2}^O a_o (y/w)^{2o}$ \\
\hline
\end{tabular}
\label{tab:keyadj}
\end{table}

When adjustment terms are used it is necessary to standardise the results to ensure that $g(0)=1$, so we can redefine the detection function as:
$$
g(y; \boldsymbol{\theta}) = \frac{k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right)}{k(0; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(0; \boldsymbol{\theta}_\text{adjust})\right)}.
$$
